---
title: "PulseR workflow"
author: "Alexey Uvarovskii"
date: '`r Sys.Date()`'
output:
  html_vignette:
    keep_md: yes
vignette: >
  %\VignetteIndexEntry{workflow}
  \VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r}
library(pulseR)
set.seed(258)
```

## Prepare a data set and conditions

### The experiment description

Let us have a labelling experiment with several time points
```{r}
replicateNum <- 3
time <- c(4,8,12)
geneNum <- 100

```

```{r echo=TRUE}
formulas <- MeanFormulas(
  total = mu,
  lab   = mu * (1 - exp(-d * time))
)
```

We create a condition matrix with two columns: the first is the condition of 
samples ("total" or "lab"), the second is the time point.
```{r echo=TRUE}
conditions <- expand.grid(condition = names(formulas), time = time)
conditions <- conditions[rep(seq_along(conditions$time),
                             each = replicateNum),]
rownames(conditions) <- paste0("sample_", seq_along(conditions$condition))
head(conditions)
```

### Generate artificial data

This section is just for reproducibility, you can easily proceed to the next one.

Create the parameters for `r geneNum` genes:
```{r}
truePar <- list(size = 1e2)
truePar$params <- data.frame(mu = 10 ^ runif(geneNum, 0, 5),
                         d = -log(runif(geneNum, .1, .9)) / 12)
rownames(truePar$params) <- paste0("gene_", 1:geneNum)

fractions <- formula(~ time + condition)
fraction_names <- levels(codeFractions(conditions, fractions))
truePar$fraction_factors <- seq_along(fraction_names)
```

Sample read counts from the negative binomial distribution:
```{r}
counts <- generateTestDataFrom(formulas, 
                               truePar,
                               conditions,
                               fractions)
```

## Create `PulseData` object
```{r}
pd <- PulseData(
    count_data = counts,
    conditions = conditions,
    formulas   = formulas,
    fractions  = ~condition+time)

```
It is important to understand, how the data are normalised.
To reduce parameter number, we use the same normalisation as in DESeq package
for samples from the *same fraction*.
Strictly, we define _fractions_ as sets of samples, which have different
amounts of RNA, estimated by the model.

### Fitting options
Now we set options for fitting. It is important to provide boundaries for the
optimal parameter search:
```{r}
opts <- setBoundaries(params = list(mu = c(.5,1e6),
                                    d = c(1e-8,1)),
                      fraction_factors = c(.1,10))
```
For other possible parameters please see "set" functions in the package
documentation (`setBoundaries, setTolerance, setFittingOptions`).  

### Initial parameter guess
Optimisation procedure may depend on the initial parameter values.
A function `initParams` provides an intyerface to simplify this step.
There are two options for how to set the parameters:

- to sample random numbers within the given boundaries 
- manual values by the user
```{r}
par <- initParameters(pulseData = pd,
                      options = opts)
```

```{r}
#par$params$mu <- guessMeans(pd)
```


## Fitting
You can specify other options as the number of cores or error 
tolerance thresholds for parameter fitting. The fitting procedure will stop,
once the difference in parameters values between two subsequent iteration is 
less than specified in `options$tolerance`.

The function `fitModel` accept the PulseData object, initial guess for the 
parameters values and fitting options.
```{r cache=TRUE}
opts$cores <- 2

opts <- setTolerance(params = 1e-3,
                     fraction_factors = 1e-2,
                     options = opts)
result <- fitModel(pd, par, opts)
```

```{r}
pr <- predictExpression(pd, result)

```


