---
title: "Formula creation"
author: "Uvarovskii Alexey"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
```

```{r echo=FALSE}
library(pulseR)

```
In the `pulseR` package, one must provide formulas which estimate the amount of 
RNA in the samples depending on the condition. 
The fitting procedure optimise parameter values in such a way, that 
the likelihood of the observed read numbers in the samples is maximal,
see [MLE](https://en.wikipedia.org/wiki/Maximum_likelihood_estimation).

There are two possibilities to define the formulas, which describe RNA
amount changes: by providing your own expressions or using implemented 
helper functions.

### Manually

Create formulas manually:

```{r}
forms <- MeanFormulas(
  total = mu,
  labelled = mu * exp(-d * time)
)
forms
```
 The formula itself is an unevaluated R expression and can be evaluated
 at given parameter values, for example:
```{r}
eval(forms$labelled, list(mu=10, d=1, time=2))
```

### Using functions

Alternatively, you may use implemented function. Let us consider a labelling
experiment.

To introduce a simple name or equation, use the function `amount`
```{r}
total <- amount("mu")
total
```
which is the same as `quote(mu)`. The unlabelled fraction is degrading during
the pulse experiment. Here we use `total` as the initial value for the
unlabelled fraction degradation:
```{r}
unlab <- degrade(total, "d", "time_L")
unlab
```
In contrast, the labelled fraction is being synthesized, and grows from the
initial zero value:
```{r}
lab <- growFrom0(total, "d", "time_L")
lab
```

If the initial value is other than zero, use the function `grow`.
For example, we make a chase experiment for time `time_C` 
after labelling for time `time_L`.
The unlabelled fraction will start to grow then from the level `unlab`:
```{r}
unlab_chase <- grow(unlab, total, "d", "time_C")
unlab_chase
```

##### Non-standard evaluation functions

Functions `grow`, `degrade` and `amount` have versions which use 
the non-standard evaluation, namely `grow_`, `degrade_` and `amount_`.
The first argument provided, must be an expression or a character string.
Other parameters, however, will be used in the form they were defined on input.

Usage example:
```{r}
degrade_(total, rate, time +1)
grow_(unlab, mu, rate, time +1)
```

