---
title: "Profile likelihood"
author: "Uvarovskii Alexey"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Confidence intervals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction



## Approximation of confidence intervals using only gene specific data

### What does it mean and when can I use it?

The profile likelihood is estimated by optimising  the whole 
likelihood function, having fixed a parameter under consideration to a 
certain value. Parameter values for different genes influence each other via 
shared parameters, for example, the normalisation factors or variance of the
negative binomial distribution.

If there are not many genes in the data, that could not be a prolblem.
However, usually experimentalists doing RNA-seq experiments are interested 
in measuring parameters of  as many genes as possible, which results in 
thousands of genes in data tables.  Optimising likelihood function for 
such cases can take a lot of computation time.
Hence it can be more efficeint to use approximate fitting by neglecting
the influence of individual gene parameters on the shared parameters.

### The workflow 

Let us use the model of experiment used in the [vignette](workflow.html)
for application  of pulseR to spike-ins-free data:

```{r include=FALSE}
library(pulseR)
```


```{r include=TRUE, cache=TRUE}
library(pulseR)
attach(pulseRSpikeinsData)
pd <- PulseData(counts, conditions, formulas, formulaIndexes, spikeins)
formulas <- MeanFormulas(
  total      =  mu,
  labelled   =  mu * exp(-d * time),
  unlabelled =  mu * (1 - exp(-d * time))
)
formulaIndexes <- list(total_fraction = "total",
                       pull_down      = c("labelled", "unlabelled"))
pd <- PulseData(counts, conditions, formulas, formulaIndexes, 
                spikeins = spikeins)
opts <- setBoundaries(list(
  mu = c(.1, 1e6),
  d = c(.01, 2)
))
opts <- setTolerance(params = 1e-3,  options = opts)
initPars <- initParameters(par = list(size=10),
                           geneParams = c("mu", "d"), 
                           pulseData  = pd, 
                           options    = opts)
fit <- fitModel(pulseData = pd, par = initPars, options = opts)
```

#### Construct the profile likelihood

```{r}
geneIndex <- 7
parName <- "d"
interval <- c(0.85,1.2) * fit[[parName]][geneIndex]
numPoints <- 51
pl <- profileGene(parName, geneIndex, pd, fit, interval, 
                  options=opts, numPoints = 21)
str(pl)
```

We can see the result using the `plotPL` function:

```{r, fig.height=4, fig.width=5}
plotPL(pl)
```

The red line corresponds to the 95% confidence level likelihood threshold.
The 95%-confidence interval for the $d$ parameter is defined from the 
crossing points of the threshold line and the profile likelihood curve.

Use the `ciGene` function in order to estimate this points numerically:

```{r}
geneIndexes <- 1:10
CIs <- ciGene(parName, geneIndexes, pd, fit, opts)
CIs
```

## Estimation of confidence intervals using the whole likelihood

If the normalisation factors are not fixed, or changes in a single gene parameter
result in observed difference in the shared or normalisation parameters values,
it is necessary to use the whole likelihood, which is a function of all shared and
individual parameters.
```{r fig.width=6, fig.height=4, eval=FALSE}
plTotal <- profile(paramPath = list("d",geneIndex),
              pd = pd, 
              par = fit,
              options = opts,
              interval = interval,
              numPoints = 20)
```

If the model includes the normalisation factors and
we are interested in estimation of the CI for the normalisation factors, 
it is possible to run the same function in the same manner:
```{r eval=FALSE}
plTotal <- profile(paramPath = list("normFactors",2,1),
              pd = pd, 
              par = fit,
              options = opts,
              interval = c(1, 3))

```
However, it may result in long computation times, since it is equivalent to
refitting the whole model several times.

```{r}
sessionInfo()
```

